An algorithm for the optimization of finite element integration

### 2016-04-07 TOMS_Paper/assembly.tex
addressing issues raised by David

% Page heads
\markboth{F. Luporini et al.}{An algorithm for the optimization of finite element integration loops}

@@

\title{An algorithm for the optimization of finite element integration loops}
\author{Fabio Luporini
\affil{Imperial College London}
David A. Ham
\affil{Imperial College London}
Paul H. J. Kelly
\affil{Imperial College London}}

@@

\acmformat{Fabio Luporini, David A. Ham, and Paul H. J. Kelly, 2016. An algorithm for the optimization of finite element integration loops.}

@@

\item For each sub-expression $\operatorname{expr}$ depending on exactly one linear loop, collect the multilinear symbols and the temporaries produced at step (1). Partition them into disjoint sets, $\mathbb{T} = \lbrace T_1, ..., T_t\rbrace$, such that $T_i$ includes all instances of a given symbol in $\operatorname{expr}$. Apply Strategy~\ref{strategy:ii} factorizing the symbols in each $T_i$, provided that this leads to a reduction in operation count; otherwise, apply Strategy~\ref{strategy:i} \\
\textit{Note: the last check ensures the flop-decreasing nature of the transformation. In the cases in which expansion outweighs code motion, Strategy~\ref{strategy:i} is preferred.}\\
\textit{Note: the expansion cost is a function of the products wrapping a symbol (how many of them and their arity), so it can be determined in a tree visit.}

@@

@@

{\color{red}FIXME: We need to put the exact versions of software in
  here. for Firedrake this should be Zenodo DOIs. This is especially true
  for the patched version of firedrake you used below}.

@@

For a fair comparison, small patches (publicly available) were written to the make \texttt{quad}, \texttt{tens}, and \texttt{ufls} compatible with Firedrake. By executing all simulations in Firedrake, we guarantee that both matrix insertion and mesh iteration have a fixed cost, independent of the transformation system employed. The patches adjust the data storage layout to what Firedrake expects (e.g., by generating an array of pointers instead of a pointer to pointers, by replacing flattened arrays with bi-dimensional ones).

@@

{\color{red}I don't think you mean ``discretized'' in the next sentence. We need to discuss what this really means.}

We discretized our problems such that the inner loop sizes would be a multiple of the machine vector length. This ensured autovectorization in the majority of code variants\footnote{We verified the vectorization of inner loops by looking at both compiler reports and assembly code.}. The biggest exception is \texttt{quad}, due to the presence of indirection arrays in the generated code. In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; the compiler reports suggest, however, that block vectorization \cite{SLP-vect} is often triggered. In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have identical structure, with loop vectorization being regularly applied.

>>>>

% Page heads
\markboth{F. Luporini et al.}{An algorithm for the optimization of finite element integration}

@@

\title{An algorithm for the optimization of finite element integration}
\author{Fabio Luporini
\affil{Imperial College London}
David A. Ham
\affil{Imperial College London}
Paul H. J. Kelly
\affil{Imperial College London}}

@@

\acmformat{Fabio Luporini, David A. Ham, and Paul H. J. Kelly, 2016. An algorithm for the optimization of finite element integration.}

@@

\item For each sub-expression $\operatorname{expr}$ depending on exactly one linear loop, collect the multilinear symbols and the temporaries produced at step (1). Partition them into disjoint sets, $\mathbb{T} = \lbrace T_1, ..., T_t\rbrace$, such that $T_i$ includes all instances of a given symbol in $\operatorname{expr}$. Apply Strategy~\ref{strategy:ii} factorizing the symbols in each $T_i$, provided that this leads to a reduction in operation count; otherwise, apply Strategy~\ref{strategy:i} \\
\textit{Note: the last check ensures the flop-decreasing nature of the transformation. In the cases in which expansion outweighs code motion, Strategy~\ref{strategy:i} is preferred.}\\
\textit{Note: the expansion cost is a function of the products wrapping a symbol (how many of them and their arity), so it can be determined through tree visits.}

@@

The Zenodo system was used to archive all packages used to perform the experiments: Firedrake \cite{lawrence_mitchell_2016_49284}, PETSc \cite{barry_smith_2016_49285}, petsc4py \cite{firedrake_2016_49283}, FIAT \cite{marie_e_rognes_2016_49280}, UFL \cite{martin_sandve_alnaes_2016_49282}, FFC \cite{anders_logg_2016_49276}, PyOP2 \cite{florian_rathgeber_2016_49281} and COFFEE \cite{fabio_luporini_2016_49279}.

@@

@@

For a fair comparison, small patches were written to the make \texttt{quad}, \texttt{tens}, and \texttt{ufls} compatible with Firedrake. By executing all simulations in Firedrake, we guarantee that both matrix insertion and mesh iteration have a fixed cost, independent of the transformation system employed. The patches adjust the data storage layout to what Firedrake expects (e.g., by generating an array of pointers instead of a pointer to pointers, by replacing flattened arrays with bi-dimensional ones).

@@

We chose the mesh dimension and the function spaces such that the inner loop sizes would always be a multiple of the machine vector length. This ensured autovectorization in the majority of code variants\footnote{We verified the vectorization of inner loops by looking at both compiler reports and assembly code.}. The biggest exception is \texttt{quad}, due to the presence of indirection arrays in the generated code. In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; the compiler reports suggest, however, that block vectorization \cite{SLP-vect} is often triggered. In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have identical structure, with loop vectorization being regularly applied.
